BIOL 274 Homework 7

### 2024-03-30 bio_dynamics/bio274_hw7.Rmd
finished biohw7

---
title: "BIOL 274 Homework 7"
author: "Ruby Krasnow"
date: "2024-03-24"
output:
  pdf_document:
    extra_dependencies: ["derivative", "tcolorbox", "xcolor"]
---
# Visualizing Solutions of Linear Systems with Constant Coefficients

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

@@

@@

@@

@@

@@

@@

\begin{tcolorbox}[colback=blue!5!white,colframe=blue!75!black,title=Check solution]
$$x'=-5C_1e^{-5 t}-C_2 e^{- t}=-4(C_1e^{-5 t}+C_2 e^{- t})+(-C_1e^{-5 t}+3C_2 e^{- t})=-4x+y $$
$$y'=5C_1e^{-5 t}-3C_2 e^{- t}=3(C_1e^{-5 t}+C_2 e^{- t})-2(-C_1e^{-5 t}+3C_2 e^{- t})=3x-2y $$
\end{tcolorbox}

@@

\end{tcolorbox}

@@

>>>>

---
title: "BIOL 274 Homework 7"
author: "Ruby Krasnow"
date: "2024-03-30"
output:
  pdf_document:
    extra_dependencies: ["derivative", "tcolorbox", "xcolor"]
---
# Visualizing Solutions of Linear Systems with Constant Coefficients

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(phaseR)
library(demodelr)
```

```{r, echo=FALSE}
# Phase Portrait setup
y0 <- matrix(c(0, 1, 0, 4, -6, 1, 5, 0.5, 0, -3), 5, 2, byrow = TRUE)

y0_df <- tibble(x=c(0,0,-6,5,0), 
                y=c(1,4,1,0.5,-3),
                IC=c("V1", "V2", "V3", "V4", "V5")) %>% 
  mutate(pt = paste0("(",x, ",", y, ")"))

fun_traj <- function(t, y, parameters) {
  x <- y[1]
  y <- y[2]
  a <- parameters[1]
  b <- parameters[2]
  c <- parameters[3]
  d <- parameters[4]
  dy <- numeric(2)
  dy[1] <- a*x + b*y
  dy[2] <- c*x + d*y  
  return(list(dy))
}

fun_plane <- c(dx ~ a*x+b*y, dy ~c*x+d*y)

make_plane <- function(params) {
  params_unnamed <- unname(params)
  traj <- trajectory(fun_traj, y0 = y0, tlim= c(0, 1), tstep=0.005,
                     parameters = params_unnamed, add=FALSE)

  t <- c(traj$t)
  x <- as_tibble(traj$x) %>% bind_cols("t"=t) %>% 
    pivot_longer(cols = c(1:5), names_to = "IC", values_to = "x")
  y <- as_tibble(traj$y) %>% bind_cols("t"=t) %>% 
    pivot_longer(cols = c(1:5), names_to = "IC", values_to = "y")
  z <- left_join(x,y, by = join_by(t, IC)) %>% filter(x<7 & x>-7 & y<7 & y>-7) %>% 
    left_join(y0_df %>% select(IC, pt), by="IC")

  p1 <- phaseplane(fun_plane, x_var="x" ,y_var="y",
                   x_window = c(-7, 7), y_window = c(-7, 7),
                   parameters = params)

  A<- matrix(params_unnamed, 2, 2, byrow=TRUE)

  lambda1 <- eigen(A)$values[1] # eigenvalue 1
  lambda2 <- eigen(A)$values[2] # eigenvalue 2

  if (abs(lambda1) < (1^(-10))) {
    v1_col <- "black"
    v1_wid <- 2
  }
  else {
    v1_col <- "gray"
    v1_wid <- 1
  }

  if  (abs(lambda2) < (1^(-10))) {
    v2_col <- "black"
    v2_wid <- 2
  }
  else {
    v2_col <- "gray"
    v2_wid <- 1
  }

  v1_slope <- eigen(A)$vectors[2,1] / eigen(A)$vectors[1,1]
  v2_slope <- eigen(A)$vectors[2,2] / eigen(A)$vectors[1,2]

  p1+theme_bw()+
    geom_abline(slope=v1_slope, color=v1_col, linewidth=v1_wid)+
    geom_abline(slope=v2_slope, color=v2_col, linewidth=v2_wid)+
    geom_path(data=z, aes(x=x, y=y, color=pt), linewidth=1.5)+
    geom_point(data=y0_df, aes(x=x,y=y, color=pt), shape=21, fill="white",size=4, stroke=2)+labs(color="IC")
}

make_plane_complex <- function(params) {
  params_unnamed <- unname(params)
  traj <- trajectory(fun_traj, y0 = y0, tlim= c(0, 1), tstep = 0.005,
                     parameters = params_unnamed, add=FALSE)

  t <- c(traj$t)
  x <- as_tibble(traj$x) %>% bind_cols("t"=t) %>% 
    pivot_longer(cols = c(1:5), names_to = "IC", values_to = "x")
  y <- as_tibble(traj$y) %>% bind_cols("t"=t) %>% 
    pivot_longer(cols = c(1:5), names_to = "IC", values_to = "y")
  z <- left_join(x,y, by = join_by(t, IC)) %>% filter(x<7 & x>-7 & y<7 & y>-7) %>% 
    left_join(y0_df %>% select(IC, pt), by="IC")

  p1 <- phaseplane(fun_plane, x_var="x" ,y_var="y",
                   x_window = c(-7, 7), y_window = c(-7, 7),
                   parameters = params)

  p1+theme_bw()+
    geom_path(data=z, aes(x=x, y=y, color=pt), linewidth=1.5)+
    geom_point(data=y0_df, aes(x=x,y=y, color=pt), shape=21, fill="white",size=4, stroke=2)+
    labs(color="IC")
}
```

@@

One positive eigenvalue and one negative eigenvalue gives a saddle:

```{r, echo=FALSE, message=FALSE, warning=FALSE,fig.keep="last"}
make_plane(c(a=1,b=2,c=3,d=2))
```

@@

Two positive eigenvalues gives a source, with solutions going towards the eigenvector $\begin{bmatrix} 2 \\ -1  \end{bmatrix}$ with the eigenvalue $\lambda=3$ of the larger magnitude.

```{r, echo=FALSE, message=FALSE, warning=FALSE,fig.keep="last"}
make_plane(c(a=4,b=2,c=-1,d=1))
```

@@

Complex eigenvalues with a positive real part gives a spiral source. Plugging in some test points (e.g., (1,1)) to our original matrix reveals that the spiral rotates counterclockwise.

```{r, echo=FALSE, message=FALSE, warning=FALSE,fig.keep="last"}
make_plane_complex(c(a=3,b=-5,c=5,d=3))
```

@@

The zero eigenvalue gives a line of equilibria, emphasized with the black line, with solutions pointing outward (because of the positive eigenvalue 10) parallel to the other eigenvector.

```{r, echo=FALSE, message=FALSE, warning=FALSE,fig.keep="last"}
make_plane(c(a=6,b=-8,c=-3,d=4))
```

@@

The positive repeated eigenvalue makes this a source, with solutions going towards the straight-line solution with the same slope (1, because y=x) as $v_1$.

```{r, echo=FALSE, message=FALSE, warning=FALSE,fig.keep="last"}
make_plane(c(a=4,b=-2,c=2,d=0))
```

@@

\begin{tcolorbox}[colback=blue!5!white,colframe=blue!75!black,title=Check solution]
$$x'=-5C_1e^{-5 t}-C_2 e^{- t}=-4(C_1e^{-5 t}+C_2 e^{- t})+(-C_1e^{-5 t}+3C_2 e^{- t})=-4x+y $$
$$y'=5C_1e^{-5 t}-3C_2 e^{- t}=3(C_1e^{-5 t}+C_2 e^{- t})-2(-C_1e^{-5 t}+3C_2 e^{- t})=3x-2y $$
\end{tcolorbox}
Two negative eigenvalues make this a sink, with solutions going towards the eigenvector $\begin{bmatrix} 1 \\ 3 \end{bmatrix}$  with the larger (less negative) eigenvalue.

```{r, echo=FALSE, message=FALSE, warning=FALSE,fig.keep="last"}
make_plane(c(a=-4,b=1,c=3,d=-2))
```

@@

\end{tcolorbox}
The zero eigenvalue gives a line of equilibria, with solutions drawn towards the line (because the other eigenvalue is negative) parallel with the other eigenvector.

```{r, echo=FALSE, message=FALSE, warning=FALSE,fig.keep="last"}
make_plane(c(a=-9,b=-3,c=3,d=1))
```

@@

Purely imaginary complex eigenvalues give a center. Testing points with our original matrix reveals a clockwise direction of rotation.

```{r, echo=FALSE, message=FALSE, warning=FALSE,fig.keep="last"}
y1<- matrix(c(0, 1, 
              10, 4,
              -6, 10,
              30, 10,
              40, 15),
          5, 2, byrow = TRUE)

y1_df <- tibble(x=c(0,10,-6,30,40), 
                y=c(1,4,10,10,15),
                IC=c("V1", "V2", "V3", "V4", "V5")) %>% 
  mutate(pt = paste0("(",x, ",", y, ")"))

traj8<- trajectory(fun_traj,
           y0         = y1,
           tlim       = c(0, 10),
           parameters = c(-3, 10, -1, 3), add = FALSE)

t8 <- c(traj8$t)
x8 <- as_tibble(traj8$x) %>% bind_cols("t"=t8) %>% 
  pivot_longer(cols = c(1:5), names_to = "IC", values_to = "x")
y8 <- as_tibble(traj8$y) %>% bind_cols("t"=t8) %>% 
  pivot_longer(cols = c(1:5), names_to = "IC", values_to = "y")
z8 <- left_join(x8,y8, by = join_by(t, IC)) %>% filter(x<50 & x>-50 & y<40 & y>-40) %>% 
  left_join(y1_df %>% select(IC, pt), by="IC")

p8 <- phaseplane(fun_plane, x_var="x" ,y_var="y",
           x_window = c(-50, 50), y_window = c(-40, 40),
           parameters = c(a=-3, b=10, c=-1, d=3))

p8+theme_bw()+
  geom_path(data=z8 %>% filter(pt!="(-6,10)"), aes(x=x, y=y, color=pt), linewidth=1.5)+
  geom_point(data=y1_df%>% filter(pt!="(-6,10)"), aes(x=x,y=y, color=pt), shape=21, fill="white",size=4, stroke=2)+labs(color="IC")
```
